Galois Characterization of Function Fields of \\ Transcendence Degree $1$ over Number Fields

### 2017-05-06 Master Thesis/introduction.tex
Add chapter overview

Given an algebraic variety $X$, we can ask ourselves how much information about $X$ is contained in the knowledge of its étale fundamental group $\pi_1^{\text{ét}}(X)$. Especially interesting are schemes which should be reconstructible solely from its étale fundamental group. If $X$ is the spectrum of a field $K$, the étale fundamental group of $X$ is just its absolute Galois group $G_K$.

@@

>>>>

Given an algebraic variety $X$, we can ask ourselves how much information about $X$ is contained in its étale fundamental group $\pi_1^{\text{ét}}(X)$. Especially interesting are schemes which are reconstructible solely from its étale fundamental group. If $X$ is the spectrum of a field $K$, the étale fundamental group of $X$ is just its absolute Galois group $G_K$.

@@

\section{Overview}

In the first chapter, we want to recap some results from valuation theory. In the second part, we will prove some lemmata in order to be able to characterize henselian fields over $K$ of valuations of rational rank $2$ in terms of their Galois cohomology groups of dimension $3$.

In the second chapter, we will show that an isomorphism $\sigma: G_K\to G_{K'}$ induces bijections on the equivalence classes of defectless valuations of $K$ and $K'$ using the cohomological characterization of henselian fields. With the Theorem of Neukirch-Uchida \ref{thm:neukirch-uchida}, we will also show that $\sigma$ induces a field isomorphism between the residue fields of corresponding defectless valuations of rational rank $1$. Using the local correspondences, we will proceed to show that $\sigma$ induces group isomorphisms $K^\times/n\to K'^\times/n$ for all $n\in\mathbb{N}$.

In the third chapter, we want to use the correspondences established in the second chapter to prove Theorem \ref{thm:main-result}. In order to do so, we need to discuss so-called free-by-finite groups. Using the existence of a so-called admissible set, we can construct a group isomorphism $K^\times\to K'^\times$. Finally, we will show that the group isomorphism can be extended to a field isomorphism $K\to K'$. By passing to the limits, we obtain a field isomorphism $\varphi: \overline{K}\to\overline{K'}$ which induces $\sigma$.

\section{Acknowledgments}

### 2017-05-06 Master Thesis/synopsis.tex
Add chapter overview

\begin{lemma}\label{spv-topology}
$\operatorname{Spv}(K)$ with the topology defined above is a quasicompact $T_0$-space. In particular the set of closed points $|\operatorname{Spv}(K)|$ of $\operatorname{Spv}(K)$ is a quasicompact $T_1$-space, endowed with the induced topology. Let $v,w\in\operatorname{Spv}(K)$. We have:
\[\mathcal{O}_w\subset\mathcal{O}_v \iff w\in\overline{\{v\}} \]
Thus $v$ is a closed point if and only if it has no non-trivial specializations.
\end{lemma}

@@

Let  $w\in |\operatorname{Spv}(K)|$. Since $\mathrm{H}^3(K_w) = \varinjlim_I \mathrm{H}^3((K_i)_w)$ there exists an $i = i(w)\in I$ such that $\operatorname{res}_{(K_i)_w/K}(g) = 0$. By Lemma \ref{2.4-closedset} there is an open neighborhood $U_{i(w)}\subset |\operatorname{Spv}(K_i)|$ of $w$ such that $\operatorname{res}_{(K_i)_v/K}(g)=0$ for all $v\in U_{i(w)}$. For $K_i\subset K_j\subset K$ let
\[ \varphi_i: |\operatorname{Spv}(K)| \to |\operatorname{Spv}(K_i)| \]
\[ \varphi_{ij}: |\operatorname{Spv}(K_j)|\to |\operatorname{Spv}(K_i)| \]
be the restriction maps. They are continuous. Since $|\operatorname{Spv}(K)|$ is quasicompact by Lemma \ref{spv-topology} we can find a finite set $S\subset |\operatorname{Spv}(K)|$ such that:
\[ \bigcup_{w\in S}\varphi^{-1}_{i(w)}(U_{i(w)}) = |\operatorname{Spv}(K)| \]
Let $j\in I$ such that $K_j$ is the composite of the fields $K_{i(w)},\ w\in S$. We also have:
\[ \bigcup_{w\in S}\varphi^{-1}_{i(w)j}(U_{i(w)}) = |\operatorname{Spv}(K_j)| \]
Let $v\in |\operatorname{Spv}(K_j)|$ and let $i = i(w),\ w\in S$ such that $v\in\varphi_{ij}^{-1}(U_{i(w)})$. We have:
\[ \operatorname{res}_{(K_j)_v/K_j}(\operatorname{res}_{K_j/K_0}(g)) = \operatorname{res}_{(K_j)_v/(K_i)_v}(\operatorname{res}_{(K_i)_v/K_0}(g)) = 0 \]
Since $K_j$ is finitely generated we obtain $\operatorname{res}_{K_j/K_0}(g) = 0$ with Lemma \ref{2.4-finitelygenerated}, thus $f = \operatorname{res}_{K/K_0}(g) = 0$.
\end{proof}

@@

Let $v\in\operatorname{Spv}_\text{d}^{(1)}(K)$ and let $N$ be the normal subgroup of $G_{K_v}$ generated by all subgroups $G_{K_w}$ for $w\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ with $v(w) = v$. Since a global field has no non-trivial algebraic extension in which every prime splits completely, we get $N=G_{K_v}$. This shows that $K'_{v'} := (K_v)'$ is a henselization of $K'$ with respect to $v'=\sigma(v)$. Furthermore, let $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ with $w_1\neq w_2$ and $v=v(w_1)=v(w_2)$, then:
\[ \sigma(G_{K^\text{t}_v}) = \sigma(G_{K_{w_1}}\cap G_{K_{w_2}}) = G_{K'_{w'_1}}\cap G_{K'_{w'_2}} = G_{(K')^\text{t}_{v'}} \]
Therefore, $\sigma$ induces with Theorem \ref{theorem-absoluteinertia} the following isomorphism:
\[ G_{k(v)} \cong G_{K_v}/G_{K_v^\text{t}} \to G_{K'_{v'}}/G_{(K')_{v'}^\text{t}}\cong G_{k(v')} \]
The rest follows from the Theorem of Neukirch-Uchida.
\end{proof}

@@

This shows that $P$ has only one pre-image under $f'$. Therefore, it has only one pre-image under $f$, say $P'$, and $P'$ has only one pre-image under $\pi$. This concludes the proof.
\end{proof}

@@

\paragraph{} Let $n\in\mathbb{N}$. In the previous remark, we have reduced the problem, so that the following diagram commutes: 
\[ \begin{tikzcd}
G_K \ar[r]\ar[d, "\sigma"'] & G_k\\
G_{K'}\ar[ru]
\end{tikzcd} \]
Now the operation of $G_K$ and $G_{K'}$ on $\mu_n$ factors through $G_k$, thus we have $\sigma(g)\xi = g\xi$ for all $\xi\in\mu_n$ and $g\in G_K$. Therefore, for all intermediate fields $L$ of $\overline{K}/K$, $\sigma$ induces an isomorphism:
\[ \varphi_n:L'^\times/n \cong\mathrm{H}^1(L', \mu_n) \to \mathrm{H}^1(L,\mu_n) \cong L^\times/n \]

>>>>

\begin{lemma}\label{spv-topology}
$\operatorname{Spv}(K)$ with the topology defined above is a quasi-compact $T_0$-space. In particular the set of closed points $|\operatorname{Spv}(K)|$ of $\operatorname{Spv}(K)$ is a quasi-compact $T_1$-space, endowed with the induced topology. Let $v,w\in\operatorname{Spv}(K)$. We have:
\[\mathcal{O}_w\subset\mathcal{O}_v \iff w\in\overline{\{v\}} \]
Thus $v$ is a closed point if and only if it has no non-trivial specializations.
\end{lemma}

@@

Let  $w\in |\operatorname{Spv}(K)|$. Since $\mathrm{H}^3(K_w) = \varinjlim_I \mathrm{H}^3((K_i)_w)$ there exists an $i = i(w)\in I$ such that $\operatorname{res}_{(K_i)_w/K}(g) = 0$. By Lemma \ref{2.4-closedset} there is an open neighborhood $U_{i(w)}\subset |\operatorname{Spv}(K_i)|$ of $w$ such that $\operatorname{res}_{(K_i)_v/K}(g)=0$ for all $v\in U_{i(w)}$. For $K_i\subset K_j\subset K$ let
\[ \varphi_i: |\operatorname{Spv}(K)| \to |\operatorname{Spv}(K_i)| \]
\[ \varphi_{ij}: |\operatorname{Spv}(K_j)|\to |\operatorname{Spv}(K_i)| \]
be the restriction maps. They are continuous. Since $|\operatorname{Spv}(K)|$ is quasi-compact by Lemma \ref{spv-topology} we can find a finite set $S\subset |\operatorname{Spv}(K)|$ such that:
\[ \bigcup_{w\in S}\varphi^{-1}_{i(w)}(U_{i(w)}) = |\operatorname{Spv}(K)| \]
Let $j\in I$ such that $K_j$ is the composite of the fields $K_{i(w)},\ w\in S$. We also have:
\[ \bigcup_{w\in S}\varphi^{-1}_{i(w)j}(U_{i(w)}) = |\operatorname{Spv}(K_j)| \]
Let $v\in |\operatorname{Spv}(K_j)|$ and let $i = i(w),\ w\in S$ such that $v\in\varphi_{ij}^{-1}(U_{i(w)})$. We have:
\[ \operatorname{res}_{(K_j)_v/K_j}(\operatorname{res}_{K_j/K_0}(g)) = \operatorname{res}_{(K_j)_v/(K_i)_v}(\operatorname{res}_{(K_i)_v/K_0}(g)) = 0 \]
Since $K_j$ is finitely generated we obtain $\operatorname{res}_{K_j/K_0}(g) = 0$ with Lemma \ref{2.4-finitelygenerated}, thus $f = \operatorname{res}_{K/K_0}(g) = 0$.
\end{proof}

@@

Let $v\in\operatorname{Spv}_\text{d}^{(1)}(K)$ and let $N$ be the normal subgroup of $G_{K_v}$ generated by all subgroups $G_{K_w}$ for $w\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ with $v(w) = v$. Since a global field has no non-trivial algebraic extension in which every prime splits completely, we get $N=G_{K_v}$. This shows that $K'_{v'} := (K_v)'$ is a henselization of $K'$ with respect to $v'=\sigma(v)$. Furthermore, let $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ with $w_1\neq w_2$ and $v=v(w_1)=v(w_2)$, then:
\[ \sigma(G_{K^\text{t}_v}) = \sigma(G_{K_{w_1}}\cap G_{K_{w_2}}) = G_{K'_{w'_1}}\cap G_{K'_{w'_2}} = G_{(K')^\text{t}_{v'}} \]
Therefore, $\sigma$ induces with Theorem \ref{theorem-absoluteinertia} the following isomorphism:
\[ G_{k(v)} \cong G_{K_v}/G_{K_v^\text{t}} \to G_{K'_{v'}}/G_{(K')_{v'}^\text{t}}\cong G_{k(v')} \]
The rest follows from the Theorem of Neukirch-Uchida \ref{thm:neukirch-uchida}.
\end{proof}

@@

This shows that $P$ has only one preimage under $f'$. Therefore, it has only one preimage under $f$, say $P'$, and $P'$ has only one preimage under $\pi$. This concludes the proof.
\end{proof}

@@

\paragraph{} Let $n\in\mathbb{N}$. In the previous remark, we have replaced $\sigma$, so that now the following diagram commutes: 
\[ \begin{tikzcd}
G_K \ar[r]\ar[d, "\sigma"'] & G_k\\
G_{K'}\ar[ru]
\end{tikzcd} \]
The operation of $G_K$ and $G_{K'}$ on $\mu_n$ factors through $G_k$, thus we have $\sigma(g)\xi = g\xi$ for all $\xi\in\mu_n$ and $g\in G_K$. Therefore, for all intermediate fields $L$ of $\overline{K}/K$, $\sigma$ induces an isomorphism:
\[ \varphi_n:L'^\times/n \cong\mathrm{H}^1(L', \mu_n) \to \mathrm{H}^1(L,\mu_n) \cong L^\times/n \]
